Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-08-30 Writing/NatureWater/1stRoundRevision/replies/reply_to_reviewers_R1.tex

\begin{itemize}
\tightlist
\item
  Data and Code Availability -- Are the data from Dr.~Haacker on the
  Github repository? Please place them there prior to publication so
  future investigators can access them rather than having to track down
  Dr.~Haacker. The data set is over seven years old so it should be made
  available according to the principles of open science.
\end{itemize}

\textcolor{blue}{The aquifer thickness data from Dr. Haacker are in the public and can be accessed from the following link: \url{https://www.hydroshare.org/resource/7d925c7944244032af98c9ed20c22db6/}. We have added this information to our data and code availability section of our revised manuscript.}

\begin{itemize}
\tightlist
\item
  Line before Equation (2) -- ``REF'\,' should be replaced with the
  actual reference.
\end{itemize}

\textcolor{blue}{The missing references have been added as follows: "[Foster et al., 2015a,b, Hrozencik et al., 2017]".}

\begin{itemize}
\tightlist
\item
  Discussion, First Paragraph -- ``reduction in aquifer depletion'' should be ``reduction in aquifer thickness''.
\end{itemize}

\textcolor{blue}{Thank you for catching this. This typo has been corrected in the revised manuscript.}

\begin{itemize}
\tightlist
\item
  Section 4.2, 3rd Paragraph -- ``the the impact'\,' -- remember to
  spell and grammar check prior to submission.
\end{itemize}

\textcolor{blue}{Thank you for highlighting this error. We have amended the text and also done a further proofread of our manuscript to capture and address and further spelling and grammatical errors.}

\begin{itemize}
\tightlist
\item
  Reference in Section 3 -- I doubt that ``Butler Jr'' is the surname, please replace with ``Butler''. Suffixes apparently aren't handled well by some of the bibliography generation programs.
\end{itemize}

\textcolor{blue}{The reference has been updated in the revised manuscript}

\begin{itemize}
\tightlist
\item
  Please remember to close all parentheses and don't mispair brackets
  and parentheses. Such stuff can raise questions about the care used in
  the analysis.
\end{itemize}

\textcolor{blue}{All parentheses and brackets have been checked and corrected where necessary in the revised manuscript}

\section{Reply to reviewer \#2}

\textcolor{blue}{Thank you for the detailed and constructive feedback on our manuscript. We have endeavored to address all your comments and suggestions, and our replies to your comments can found in blue below (original comments in black).}

This is an interesting paper that asks an important yet answered
question in the field of agricultural sustainability - how much is
groundwater depletion reducing agricultural productivity in the High
Plains region in the United States? While the paper is an important
contribution to the literature, there are several issues that need to be
addressed prior to acceptance at any journal.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  There needs to be more description about why aquifer saturation
  thickness is an appropriate proxy for farmers' ability to irrigate
  from wells. This is particularly true since this is a broad interest
  journal, and the mechanism for the main independent variable should be
  very clear to all readers.
\end{enumerate}

\textcolor{blue}{Aquifer thickness affects farmers ability to meet the water requirements of their crops due to its impacts on the pumping yields (or capacities) of groundwater irrigation wells. As the thickness of an aquifer declines, the rate at which water can flow from the aquifer to the pumping well is reduced proportionally to the aquifer's transmissivity (equal to the hydraulic conductivity of the aquifer geology multiplied by the saturated thickness of that aquifer geology). For small relative changes in aquifer thickness (e.g. for initial depletion of an aquifer with large initial thickness), the change in well yield is small and unlikely to have substantial negative impacts on drought risk exposure as a farmer could simply run their pump for slightly longer to compensate for the small reduction in instantaneous well yield. However, for larger relative changes in aquifer thickness (e.g. due to significant aquifer depletion and/or thinner initial aquifer bodies), the magnitude of well yield reductions will be much larger. These reductions in water supply capacity cannot be mitigated by changes to irrigation scheduling or management practices. The farmer will often be unable to keep up with crop water demands during the peak of the growing season in the absence of rainfall, even if running their system continually. This effect can also typically not be mitigated by drilling additional or deeper wells, due to a combination of both regulatory moratoriums on well drilling - which are common across the HPA - and high costs of deep well construction. To more clearly highlight for non-specialist readers these connections between saturated thickness, well yields, and drought resilience of agriculture, we have added the following text on page 1 lines 40-47 of our revised manuscript. We note that we are limited in elaborating further on background information given journal word constraints, and hence provide key supporting studies in citations for readers interested in deepening understanding of these dynamics.}

\textcolor{blue}{"...depletion also reduces the thickness of an aquifer, lowering the aquifer's transmissivity (i.e., the potential rate of water supply to wells). This latter effect reduces the rate at which water can be pumped and applied to a field as irrigation (Konikow and Kendy, 2005; Foster et al., 2015; Hrozencik et al., 2017). If the reduction in well yields is substantial - for example as would occur when a large proportion of aquifer thickness is depleted (Hecox et al., 2002; Korus and Hensen, 2020) - then a farmer may be unable to extract sufficient water to meet crop water demands, in particular during periods of limited rainfall and in the peak of the growing season when crops are most sensitive to water deficits (Foster et al., 2015a; Rouhi-Rad et al., 2020)."
}

\textcolor{blue}{Additional references added:}

\begin{itemize}
\item
  \textcolor{blue}{Hecox, G.R., Macfarlane, P.A., and Wilson B.B. (2002). Calculation of yield for High Plains wells: relationship between saturated thickness and well yield. Open file report 2002–25C, Kansas Geological Survey: Lawrence, Kansas, USA.}
\item
  \textcolor{blue}{Korus, J. T., and Hensen, H. J. (2020). Depletion percentage and nonlinear transmissivity as design criteria for groundwater-level observation networks. Environmental Earth Sciences, 79(16), 382.}
\item
  \textcolor{blue}{Rouhi-Rad, M., Araya, A., and Zambreski, Z. T. (2020). Downside risk of aquifer depletion. Irrigation Science, 38, 577-591.}
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Though I am not trained as a hydrologist, my understanding is that
  aquifer saturation thickness is a good measure of general water
  availability in an aquifer but not necessarily how much water is
  available at a specific location or point. If this is the case, the
  authors should make clear why this is an appropriate measure to use in
  their study and potential limitations of using this approach.
\end{enumerate}

\textcolor{blue}{Aquifer thickness is a spatially and temporally variable property. It  represents the variation in water storage at specific places in an aquifer and at specific points in time, not the overall aggregated storage within an aquifer. Aquifer thickness also influences the transmissivity of an aquifer, which in turn is proportional to well yields. Aquifer thickness therefore influences the potential rate of water extraction at different points within an aquifer and how this varies over time (e.g., due to depletion reducing aquifer thickness). As noted in our reply to another reviewer, it is important to also note that the relationship between aquifer thickness and well yields will also be influenced by spatial variations in aquifer geology (specifically hydraulic conductivity which together with saturated thickness defines aquifer transmissivity). Data on county-level spatial variations in aquifer hydraulic conductivity is not available for HPA (or indeed for other aquifers in general), meaning we cannot control for these variations thus adding noise or uncertainy to our analyses. This noise is referred to as measurement error in econometric studies, and can result in attenuation bias. Specifically, measurement error in an explanatory variable (here, aquifer thickness) tends to bias your estimation in a way that makes the impact of the explanatory variable with measurement errors appear "less" influential than it truly is. So, in our context, our estimated impact of aquifer thickness on the share of irrigation is conservative because measurement error in the relationship between aquifer thickness and well yields will be attenuating some of the true impact of aquifer thickness on drought risk. To highlight these points, we have added the following texts in the Discussion section on page 10 lines 210-224 of our revised manuscript.}

\textcolor{blue}{"Unobserved variations in aquifer properties will also introduce noise into the relationships between aquifer thickness, well yields, and drought risk. For example, a thinner aquifer interval composed of coarse sands and gravels may be able to supply comparable or higher well yields than a thicker aquifer dominated by fine sands and silt deposits, and hence offer greater resilience to water deficits despite similar or larger aquifer thickness [Butler et al., 2013, Korus and Hensen, 2020]. These measurement errors in the relationship between aquifer thickness and well yield will lead to attenuation bias [Bound and Krueger, 1991, Hyslop and Imbens, 2001], suggesting that our analysis is likely to be an underestimate of the true impacts of groundwater depletion on drought risk. Nonetheless, our analysis shows that - even in the presence of these measurement uncertainties - we are still able to identify statistically significant increases in drought risk exposure for irrigated agriculture as a function of declining saturated thickness. Improved data on spatial variations in aquifer properties and greater monitoring of real-world changes in well capacities alongside traditional water level measurements would help to further refine understanding of patterns of drought vulnerability across the HPA. This would enable more precise estimates to be made of critical tipping points in the resilience of irrigated agriculture to drought, which are needed to effectively design and target groundwater conservation policies."}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  It is unclear to me why county and time fixed effects were not used in
  every regression. In particular, it does not look like they were
  included in the regression that modeled irrigated area as a function
  of saturated thickness and water deficit. Yet, accounting for
  time-invariant differences across counties is important, especially
  because this is the regression that showed the largest effect of
  saturated thickness on a measure of agricultural production (irrigated
  area). Is there not enough variation within a county through time to
  pick up the effect of saturated thickness on irrigated area? This
  should all be made clearer, and a model that better captures the
  causal relationship between saturated thickness and area irrigated
  should be used if possible. If it's not possible, the limitations of
  the current approach should be made very clear and potential issues
  with inference.
\end{enumerate}

\textcolor{blue}{Thank you for this comment, which we believe is an important point to clarify in terms of our methods and the assumptions made in our analysis. County and year fixed effects were included in equation (1), but only year fixed effects were included in equation (2). As you guessed, the problem of including county FEs lies in the fact that this would eliminate a large proportion of the variation in aquifer thickness, which is predominantly cross-sectional rather than longitudinal. To illustrate this, Figure 6 in Appendix A provides histograms of (1) aquifer thickness demeaned by the county average, (2) aquifer thickness demeaned by the state average, and (3) aquifer thickness without demeaning. As shown in Figure 6, inclusion of county FEs (mathematically equivalent to regression using county-demeaned aquifer thickness) dramatically reduces the total variation in aquifer thickness leaving little variations to accurately estimate the impact of aquifer thickness on the share of irrigated acres. The inclusion of State FEs does not diminish the variation in aquifer thickness. Therefore, while the inclusion of State FEs does not provide as tight of a control as county FEs, we have decided to include State FEs for both corn and soybean irrigated share regressions to provide a balance between preserving spatial variations in aquifer thickness and controlling for other spatially dependent effects. Figure 7 compares the estimated impacts of aquifer thickness with or without State FEs. The negative impact of reduction in aquifer thickness on the share of irrigated acres is slightly greater when state fixed effects are included. This suggests that aquifer thickness is unlikely to be very highly correlated with important unobservable variables that are affecting irrigated share significantly. Given the direction of potential bias, providing tighter controls through inclusion of county fixed effects therefore may result in even more significant drop in the share of irrigated acres with smaller aquifer thickness suggesting our results are possibly a conservative estimate of the true impact of aquifer thickness on drought risk. A detailed discussion of these analyses and tradeoffs is now provided in Appendix A.
}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  There are minor typos (e.g., missing reference listed as REF) that
  should be corrected prior to resubmission.
\end{enumerate}

\textcolor{blue}{The missing in-text citation has been corrected, and the revised manuscript has been thoroughly proofread to catch any remaining spelling, grammatical, and formatting errors.}

\section{Reply to reviewer \#3}

The paper aims to show the impacts of aquifer depletion on irrigated
crop yields and production areas in the United States High Plains. To do
so, the paper applies a panel data analysis approach to a detailed
county-level dataset of saturated thickness, water deficit, yields, and
land-use, using a spline model. The authors find that aquifer depletion
reduces farmers' ability to cope with water deficits.

The significance and novelty of the paper stems from the clear and
encompassing empirical approach to the subject matter. The depletion of
the Ogallala aquifer and implications for irrigated agriculture are not
new per se. A compelling quality of the paper is the clear statistical
evaluation of a large panel dataset, constructed to show the negative
impacts of aquifer depletion on farmer's ability to cope with water
deficits, over a long period of time. To fully harness the potential of
this quality for readers, the authors should mainly consider improving
the clarity of the figures presented and of their description in the
text, and addressing some other points that remained unclear (see
details below).

\textcolor{blue}{Thank you for the detailed and constructive feedback on our manuscript. We have endeavored to address all your comments and suggestions, and our replies to your comments can found in blue below (original comments in black).}

\hypertarget{main-points}{%
\section{MAIN POINTS:}\label{main-points}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Abstract:
\end{enumerate}

\begin{itemize}
\tightlist
\item
  ``We show that aquifer depletion reduces the ability of farmers to
  sustain irrigated crop yields and production areas in drought years.''
  - If stated this way, please clarify in the results and/or discussion,
  how causality (direction of causality, no omitted factors) is
  established.
\end{itemize}

\textcolor{blue}{Thank you for raising this important point - we agree that clarifying the extent to which we can infer causality from our analysis is important. We discuss the underlying direction of causality in detail in the main section of our analysis, highlighting the biophysical mechanisms through saturated thickness of an aquifer influences well yields and how well yields will in turn affect drought risk exposure for a farmer. This discussion has been expanded on page 1 lines 37-54 to provide more detail on the physical mechanisms, and supporting empirical and theoretical evidence, connecting saturated thickness, well yields, and water-related crop production risks. For these reasons, we believe that our results and interpretation of our findings are strongly supported by and consistent with hydrologic and economic theories, and provide useful insights for groundwater management even if causality cannot be formally determined.}

\textcolor{blue}{We acknowledge that a further important factor when seeking to establish causality is the issue of possible bias in the estimation of the impact of a variable interest - in our case the effects of saturated thickness on agricultural drought risk. We do not intend to claim that our estimation is free of bias. Indeed, in all empirical studies, it is almost impossible to have no unobserved (omitted) factors unless the data is from a randomized control trial. Other sources of bias (e.g., model misspecification error) will also almost always exist. In our analysis, one of the main limitations is not including county fixed effects in our regressions, which would have controlled all the county-specific time-invariant characteristics. County-level fixed effects were not included because the majority of the variation in saturated thickness is cross-sectional rather than time-dependent, and incorporating county fixed effects would therefore have removed this variation. We acknowledge this as a limitation, and have added a detailed discussion in both the Methods (page 15 lines 339-344) and Appendix C of our revised manuscript.}

\textcolor{blue}{Furthermore, our analysis will also be influenced by measurement errors in aquifer thickness and its relationship with  well yield (e.g., due to spatially variable hydraulic conductivity of aquifer layers). This limitation is highlighted (with expanded text compared to our original manuscript) in the Discussion section (page 10 lines 210-220). Importantly, these measurement errors would be expected to result in attenuation bias, making our estimates of the impact of aquifer thickness smaller in magnitude than the likely true effect.}

\begin{itemize}
\tightlist
\item
  ``\ldots{} in drought years'' - Please define in the in the main text
  (or methods) how drought and drought years are defined. E.g. are any
  effects linked to increasing water deficits considered drought-related
  or only those above a certain level?
\end{itemize}

\textcolor{blue}{In this paper, we do not seek to formally define a threshold level of water deficit beyond which a year is considered to be a `drought'. Instead, we define meteorological conditions during the growing season in terms of a common water deficit metric - which provides a continuous measure of how `drought' a given year was in a specific county considering both precipitation and crop evapotranspiration demands. A water deficit of 1000 mm, for example, could be considered to be a "severe" drought as precipitation is substantially less than rainfall with the result that corn and soybean yields are estimated to be almost zero under rainfed production. In contrast, a water deficit of -200 mm would be considered not to be a drought because precipitation exceeds crop evapotranspiration, and as a result irrigated and rainfed yields are almost identical. We do agree that the wording "in drought years" may be somewhat confusing in this context given the lack of a formal threshold beyond which a year is considered to be drought. Therefore, we have amended this phrase throughout the manuscript to instead say "in years with large water deficits".}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Figure 1: This figure does not convey the effects of saturated
  thickness very clearly, either because the underlying analysis does
  not produce substantial effects, or because they are not presented in
  a compelling manner. It might also be helpful to use a visualization
  that more effectively illustrates the statements in the text. See
  detailed points below:
\end{enumerate}

\textcolor{blue}{We appreciate this comment. In response, we have added an additional figure - Figure 2 in the revised manuscript - which shows in more detail the difference in yield between the 2nd (medium level of saturated thickness) and 3rd (highest level of saturated thickness) quantiles relative to the 1st quantile (lowest level of saturated thickness). This illustrates the differences in irrigated corn and soybean yields across different aquifer thickness levels. Please see our detailed responses to your comments below for more explanation of the interpretation of this figure with regard to distinguishing significant and non-significant results from our analysis.}

\begin{itemize}
\tightlist
\item
  Please distinguish more clearly between significant and
  non-significant results. P. 2 states that ``counties with the highest
  level of saturated thickness \ldots{} experience \ldots{} productivity
  close to rainfed yields''. From the figure it is difficult to tell if
  the 2nd and 3rd quantile are significantly different for soybean. For
  corn this does not seem to be the case. Similarly, the text at the
  bottom of p.~2 states that ``saturated thickness has a statistically
  significant impact on per-area irrigated crop yields in moderate to
  extreme drought conditions''. Please clarify how the significance is
  evaluated here. Which range of deficits is meant by ``moderate to
  extreme''? Are the soybean yields considered significantly different
  between quantiles? Are only the 1st and 3rd quantile considered
  significantly different for corn? - It could even be of interest to
  readers which results are not significant here, in contrast to the
  greater significance of the effects analyzed in Fig. 2 and Fig. 3.
\end{itemize}

\textcolor{blue}{We have broken the above comment into separate points, which we answer  individually below.}

Your comment: Please distinguish more clearly between significant and
non-significant results.
